---
title: "Epidemiological Methods I"
author: "Jacob Simmering, PhD"
date: "February 26, 2018"
output:
  beamer_presentation: default
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE,
                      warning = FALSE)
library(tidyverse)
library(rdrobust)
```

## Objectives

- Describe the major methods and approaches used in observational data analysis
- Describe how these methods differ from both experimental and quasi-experimental approaches
- Describe when to use a cohort study, a case-control study and other designs
- Describe the advantages and disadvantages of each design

## Next Week

- Bias
- Confounding
- Threats to validity

## Review of Experimental Methods

The causal effect of a treatment or exposure is of primary interest to us in 
most of our research projects 

Does the treatment reduce the intensity of disease? Does the exposure increase
risk of a certain outcome?

## Review of Experimental Methods

The causal effect of a treatment or exposure is of primary interest to us in 
most of our research projects 

Does the treatment reduce the intensity of disease? Does the exposure increase
risk of a certain outcome?

In other words:
$$E(Y | T = 1) - E(Y | T = 0)$$

## Fundamental Problem of Causal Inference

With a given subject $s$, we can only observe one outcome:

Either you are or are not treated 

Either you are or are not exposed

If we can only observe one outcome per subject, how can we estimate the effect
of a variable?

## Randomization

If we randomly assign people to be treated or not treated (exposed or 
not exposed), then we can easily estimate the causal difference. 

On average, the only difference between the groups is the treatment, which was 
randomly assigned, and so the differences in the outcome in the two groups is 
the same as the causal effect of the variable that we randomized on.

This is a very powerful method for assessing the effect of some variable but it
is not always suitable. 

## Randomization is Not Always Possible

While randomization may be a "gold standard" for showing causal effects, it is 
not always possible for both ethical and logistical reason

Note that the The "gold standard" claim of RCTs have been challenged by some, 
see Deaton A, Cartwright N, "Understanding and misunderstanding randomized 
controlled trials" in *Social Science & Medicine*, 2017

## Randomization is Not Always Ethical
![](/Users/jsimmeri/projects/epi_methods/bmj_parachute_rct.tiff)

## Randomization is Not Always Practical

Randomization is poorly suited to questions about 

- effects with long latent periods
- rare outcomes

due to the massive scale required to follow up and maintain a sample size of 
sufficient power

## Randomization is Not Always Desirable

Randomized trials have (relatively) high costs and so generally the population
in the trial is selected to maximize the trials ability to find a significant
effect at the lowest possible sample size, time and cost

As a result, complicated patients, old patients and others are frequently 
excluded from these trials

Internal validity is optimized by randomized, but external validity and 
applicability is sometimes unclear

## Randomization is Not Always Desirable

What is the validity of an estimate about a drug for treating diabetes in a 
trial that excluded anyone with heart disease given that many people with 
diabetes also have heart disease?

Real world treatment effects may differ significantly from the treatment 
effects observed in controlled trials

## Inference Without Randomization

Without random assignment, there is concern that the assignment process may be
related to the expected outcome or both to the exposure and outcome

- A study of coffee drinking and cancer may be confounded by the fact many 
people (used to) smoke while drinking coffee

- A study of the effect of a headache medication on headache intensity after 
60 minutes may be confounded if people who sleep less are more likely to get 
headaches and also more likely to medicate them

## All Is Not Lost

Without randomization, different study designs are called for to manage and 
characterize observational research

The two major designs are

- cohort studies
- case-control studies

We will cover these and several less common designs in detail today

## Cohort Studies

In a nutshell, cohort studies

- recruit based on exposure status
- determine outcome status
- good for rare exposures
- bad for rare diseases

## Cohort Studies

Exposure or treatment is known as the time of enrollment and subjects can be 
selected specifically for their exposure/treatment status

At the time of the enrollment, the outcome state for a subject may or may not 
be known

## Cohort Studies

Studies where the outcome is not yet realized at the time of enrollment but 
subjects are followed for the development of the outcome are considered 
prospective studies

Studies where the outcome is realized at the time of enrollment are considered
retrospective studies

## Cohort Studies

In general, prospective studies are considered to have greater validity as they
avoid potential issues with temporal misordering of exposure and outcome 
onset but take longer to conduct

## Relative Risk

The relative risk is the primary measure of association in cohort studies and
compares the risk of developing the disease/outcome of interest in the exposed
and unexposed cohorts

$$RR = \frac{P(O = 1 | E = 1)}{P(O = 1 | E = 0)}$$

## Relative Risk

The relative risk is the primary measure of association in cohort studies and
compares the risk of developing the disease/outcome of interest in the exposed
and unexposed cohorts

$$RR = \frac{\frac{N(O = 1, E = 1)}{N(E = 1)}}{\frac{N(O = 0, E = 0)}{N(E = 0)}}$$

## Relative Risk

With the standard 2x2 table:

|             | Diseased | Not Diseased |
|-------------|----------|--------------|
| Exposed     |     a    |      b       |  
| Not Exposed |     c    |      d       |

$$RR = \frac{a/(a + b)}{c/(c + d)}$$

## When to Use a Cohort Study

- Rare exposure
- Want to directly estimate risk/relative risk
- Want to collect data prospectively

## When Not To Use a Cohort Study

- Rare diseases

## Case-Control Studies

In a nutshell, case-control studies

- recruit based on outcome status
- determine exposure status
- good for rare diseases
- bad for rare exposures

## Case-Control Studies

Outcome status is known at the time of enrollment and subjects are selected 
specifically for their outcome status

Exposure status is then determined

## Case-Control Studies

All case-control studies are retrospective since the outcome must be realized
at the time of enrollment

There is concern about the potential for different rates of recall (or 
misrecall) of exposures by case/control status

There are greater issues with determining temporal ordering than in prospective
cohort studies

## Case-Control Studies Studies

In general, case-control studies are considered to be more sensitive to bias
and confounding than cohort studies

Awareness of this potential, the nature of your data and the statistical methods
you use can manage this risk

## Odds Ratio

Since the rate of disease in a case-control study is determined by the 
researcher, relative risks cannot be calculated

However, the odds ratio is not sensitive is insensitive to the type of sampling
and therefore can be used in case-control studies

## Odds

The odds of an event are related to the probability of an event $E$ as:

$$\text{Odds}(E) = \frac{P(E)}{1-P(E)}$$

Typically this is expressed as "X to Y for/against"

## Odds

An event with $P(E) = 0.25$ is 1 to 3

An event with $P(E) = 0.75$ to 3 to 1

## Odds Ratio

The odds ratio is the ratio of the odds of having the outcome of interest in the
exposed and unexposed subjects in the case-control study

$$OR = \frac{\text{Odds}_{E = 1}}{\text{Odds}_{E = 0}}$$


## Odds Ratio

The odds ratio is the ratio of the odds of having the outcome of interest in the
exposed and unexposed subjects in the case-control study

$$OR = \frac{N(E = 1, O = 1) / N(E = 1, O = 0)}{N(E = 0, O = 1) / N(E = 0, O = 0)}$$

## Odds Ratio

The odds ratio is the ratio of the odds of having the outcome of interest in the
exposed and unexposed subjects in the case-control study

$$OR = \frac{N(E = 1, O = 1)N(E = 0, O = 0)}{N(E = 0, O = 1)N(E = 1, O = 0)}$$

## Odds Ratio

When using the standard 2x2 table:

|             | Diseased | Not Diseased |
|-------------|----------|--------------|
| Exposed     |     a    |      b       |  
| Not Exposed |     c    |      d       |

$$OR = \frac{a / b}{c / d} = \frac{ad}{bc}$$

## Odds Ratio

Compared to the probability and relative risk discussed in the cohort study,
the odds and odds ratios are actually rather unintuitive

It is important to note that an OR does not describe changes in risk (which 
typically implies probability) but rather changes in the odds of the outcome

## Rare Disease Assumption

Under certain circumstances, we note that $$RR \simeq OR$$

Specifically, when the incidence of the disease is relatively rare in the 
population

This is known as the rare disease assumption

## Rare Disease Assumption

If it the case that $N(E = 1, O = 1) \gg\ N(E = 1, O = 0)$, then

- $$N(E = 1, O = 0) \simeq N(E = 1, O = 0) + N(E = 1, O = 1)$$
- $$N(E = 0, O = 0) \simeq N(E = 0, O = 0) + N(E = 0, O = 1)$$

## Rare Disease Assumption

$$RR = \frac{N(E = 1, O = 1) / N(E = 1)}{N(E = 0, O = 1) / N(E = 0)} \simeq$$
$$\frac{N(E = 1, O = 1) / N(E = 1, O = 0)}{N(E = 0, O = 1) / N(E = 0, D = 0)} = OR$$

## Rare Disease Assumption

When the disease is rare, the $RR \approx OR$

How rare is rare? Generally, under 5-10% of the population results in a 
reasonably small bias in treating the OR like the RR

## Rare Disease Assumption

```{r}
tibble(
  p = rep(seq(0.001, 0.40, 0.001), 5),
  rr = rep(c(0.5, 0.9, 1.1, 1.5, 2), each = 400)
) %>%
  mutate(a = (p * rr) / (rr + 7 / 3),
         c = p - a, 
         b = 0.3 - a,
         d = 0.7 - c) %>%
  mutate(or = (a * d) / (b * c)) %>%
  mutate(RR = as.factor(rr)) %>%
  ggplot(aes(x = p * 100, y = or, color = RR)) + 
  geom_line() + 
  labs(x = "Percent of population with disease",
       y = "Estimated OR")
```

## When to Use a Case-Control Study

- Rare diseases and outcomes
- Costly measurements required (e.g., biomarkers)

## When not to Use a Case-Control Study

- Rare exposures

## Cohort and Case-Control Summary

|            | Cohort Studies   | Case Control Studies | 
|------------|------------------|----------------------|
| Grouped by | Exposure Status  | Outcome Status       |
| Discovered | Outcome Status   | Exposure Status      |
| Measure    | Relative Risk    | Odds Ratio           |
| Useful for | Rare Exposures   | Rare Diseases        |
| Poor for   | Rare Diseases    | Rare Exposures       |
| Timeline   | Both pro and retrospective | Only retrospective |

## Other Commonly Used Designs
While cohort and case-control studies are used for most observational research,
other designs may be more useful depending on the question. Some of these 
designs are

- Nested case-control
- Case crossover
- Case-time-control
- Cross sectional
- Ecological
- Pre/Post
- Interrupted time series and regression discontinuity

## Nested Case-Control
Nested case-control studies are case-control studies done where the cases and
controls are sampled from a larger cohort or similar study

The advantages of such a design include easily identifying potential subjects

This type of design is most useful when you are doing a cohort study and are 
interested in a secondary question, especially if the exposure of interest 
requires significant investment in time or money to measure 

Measuring the exposure on controls exhibits diminishing marginal returns of 
information

## Case-Crossover
Case-crossover is useful when studying transient and short exposures and acute, 
rapidly detected and short-lived outcomes

Frequently used when studying the effects of weather on health 

Controls are generated by using the case by offset by a some time

## Case-Crossover

For instance, if you were interested in the effects of weather on the risk of
the flu, you might find all cases of the flu and then look at the weather 
immediately before the subject got the flu and then compare that to the weather
immediately before the same date last year, two years ago and so on when the 
case did not develop the flu

The design implicitly matches on seasonal factors and as the case serves also 
serves as the control many subject factors are also controlled

## Case-Time-Control
(Diff in diff?)

Describe the differences between a control group and an exposed group over time,
especially after an intervention or change in policy

## Case-Time-Control (Graphically)
```{r}
set.seed(222)
did <- tibble(
  state = rep(c("IA", "MN"), each = 104),
  t = rep(1:104, 2),
  drug_use_high = rnorm(104 * 2, 40, 10),
  drug_use_low = rnorm(104 * 2, 30, 10),
  drug_use_observed = ifelse(state == "IA" & t >= 53, drug_use_low,
                             drug_use_high)
) 

ggplot(did, aes(x = 2016 + (t - 1) / 52, y = drug_use_observed, 
             color = state)) + 
  geom_line() + 
  labs(x = "Year", y = "Weekly Usage Rates")
```

## Case-Time-Control (Graphically)
```{r}
did %>%
  mutate(group = ifelse(t >= 53, paste(state, "After"),
                        paste(state, "Before"))) %>%
  ggplot(aes(x = 2016 + (t - 1) / 52, y = drug_use_observed, 
             color = state)) + 
  #geom_line() + 
  geom_smooth(method = "lm", alpha = 0, formula = y ~ x + I(x >= 2017)) + 
  labs(x = "Year", y = "Weekly Usage Rates")
```

## Cross-Sectional
Current and immediate levels of exposure and disease burden

Relatively weak form for causal inference but may suggest future research 
questions

## Ecological
Similar to a cross-sectional design but with the data aggregated 

Correlations between two variables summarized at some unit level (state, county,
etc)

Potentially misleading summary measures are possible

## Gender Preference in Admissions at UCB in 1973
```{r}
UCBAdmissions %>%
  as_tibble() %>%
  mutate(n = as.integer(n)) %>%
  select(Dept, Gender, Admit, n) %>%
  group_by(Gender, Admit) %>% 
  summarize(n = sum(n)) %>%
  spread(Admit, n) %>%
  mutate(pct_admitted = Admitted / (Admitted + Rejected)) %>%
  ggplot(aes(x = Gender, y = pct_admitted * 100)) + 
  geom_bar(stat = "identity") + 
  labs(x = "Gender", y = "Admission Rate (Percent)") + 
  ggtitle("Admission Probability by Gender at UCB 1973")
```

## Gender Preferences in Admissions at UCB in 1973
```{r}
UCBAdmissions %>%
  as_tibble() %>%
  mutate(n = as.integer(n)) %>%
  select(Dept, Gender, Admit, n) %>%
  group_by(Gender, Dept, Admit) %>% 
  summarize(n = sum(n)) %>%
  spread(Admit, n) %>%
  mutate(pct_admitted = Admitted / (Admitted + Rejected)) %>%
  ggplot(aes(x = Gender, y = pct_admitted * 100)) + 
  geom_bar(stat = "identity") + 
  labs(x = "Gender", y = "Admission Rate (Percent)") + 
  facet_wrap(~ Dept) + 
  ggtitle("Admission Probability by Gender and Department at UCB 1973")
```

## Gender Preferences in Admissions at UCB in 1973
Unadjusted, pooled analysis suggested men had odds of admission that were 1.8
times greater than those of female applicants

On a department-by-department basis, there is no evidence of preference for 
males - indeed the only unbalanced department actually shows a preference for 
females

Female applicants were more likely to apply to highly selective and competitive 
departments whereas male applicants were more likely to apply to departments 
that were less competitive and tended to take many/most well-qualified 
applicants

## Ecological
Ecological studies that use summary measures may provide misleading results as
the summary measures may be misleading due to confounding

Poor form of showing causality, potentially hypothesis generating

## Pre/Post
Measure the outcome of interest before and after an intervention and compare
the values

May be done without a control group but at risk to bias due to regression to the
mean and failure to control for natural, potentially pre-existing processes

Ideally, you would want a series of several data points on either side of the 
intervention to have an assessment of how variable the data generating process 
is and a control group as a measure of any other effects that may be occurring

## Interrupted Time Series/Regression Discontinuity 
```{r}
data("Seatbelts")

Seatbelts %>%
  as_tibble() %>%
  mutate(date = time(Seatbelts)) %>%
  ggplot(aes(x = date, y = (DriversKilled + front + rear + VanKilled) / kms * 1000)) + 
  geom_line() + 
  annotate("rect",
           xmin = 1983.083,
           xmax = 1985,
           ymin = 0,
           ymax = Inf,
           alpha = 0.25) + 
  labs(x = "Date (Shaded portion has law in effect)",
       y = "Number of Traffic Deaths Per 10,00 km driven")
```


## Interupted Time Series

```{r}
Seatbelts %>%
  as_tibble() %>%
  mutate(date = time(Seatbelts)) %>%
  ggplot(aes(x = date, y = (DriversKilled + front + rear + VanKilled) / kms * 1000)) + 
  geom_line() + 
  geom_smooth(aes(group = law), method = "lm", alpha = 0) + 
  annotate("rect",
           xmin = 1983.083,
           xmax = 1985,
           ymin = 0,
           ymax = Inf,
           alpha = 0.25) + 
  labs(x = "Date (Shaded portion has law in effect)",
       y = "Number of Traffic Deaths Per 10,00 km driven")
```

## Regression Discontinuity 
Imagine there is some variable that has some threshold:

- Blood pressure over 140/90
- Admitted to school
- Treated

And you want to know what is the causal effect of being
above the threshold compared to being below

## Regression Discontinuity
```{r}
data("rdrobust_RDsenate")

rdrobust_RDsenate %>%
  as_tibble() %>%
  ggplot(aes(x = margin, y = vote, group = margin > 0)) + 
  geom_point() + 
  coord_cartesian(xlim = c(-10, 10)) + 
  #geom_smooth(method = "lm", alpha = 0) + 
  labs(x = "Margin of Democrat Win/Loss in Prior Election",
       y = "Democrat Vote Share in Current Election")
```

## Regression Discontinuity

Fit a regression model to the data on either side of the cutoff $c$
and compare the difference in the fitted values as the models 
approach $c$

## Regression Discontinuity
```{r}
rdrobust_RDsenate %>%
  as_tibble() %>%
  ggplot(aes(x = margin, y = vote, group = margin > 0)) + 
  geom_point() + 
  coord_cartesian(xlim = c(-10, 10)) + 
  geom_smooth(method = "lm", alpha = 0) + 
  labs(x = "Margin of Democrat Win/Loss in Prior Election",
       y = "Democrat Vote Share in Current Election")
```

## Regression Discontinuity
Two major considerations

- Is it truly the case that those at $c+\epsilon$ and $c-\epsilon$ are similar?
- How much data should I use on either side of the cutoff to fit the local 
models?